Factor-specific variances in R
========================================================

Under construction!

```{r echo=FALSE}
opts_chunk$set(fig.width=4,fig.height=4)
```

A topic that has come up several times on the R list and elsewhere: (how) can one allow different variances among random effects within different categories?  It turns out that this morphs into a couple of other topics of interest: narrowly, how can one fit a random effects model with diagonal variance-covariance model within a grouping factor?  More generally, how can one fit random effects model with different variance-covariance structures in the random effects?

## Easy example: factor-specific *residual* variances

We don't even need a mixed model for this one, we can just use generalized least squares (`gls()` in the `nlme` package).  Suppose
$$
\begin{split}
Y_{ij} & = \delta_i + \epsilon_{ij} \\
\epsilon_{ij} & \sim \text{Normal}(0,\sigma^2_i) ;
\end{split}
$$
that is, the residual variance differs by group.

```{r simex1}
set.seed(101)
dsd <- c(1,2,3)
d <- expand.grid(f=factor(letters[1:3]),rep=1:100)
delta <- 4:6
d$y <- rnorm(nrow(d),mean=delta[d$f],sd=dsd[d$f])
library("ggplot2")
theme_set(theme_bw())
qplot(f,y,geom="boxplot",data=d)
```

```{r gls1}
library("nlme")
m1 <- gls(y~f-1,data=d,weights=varIdent(form=~1|f))
```

We could look at the whole thing, but restricting ourselves for brevity to the fixed-effect parameter estimates
and the estimated variances (parameterized as ratios of residual variances in groups $2 \dots n$ to the variance in the first group):
```{r glsresults}
coef(m1)
summary(m1$modelStruct$varStruct)
```

The linear mixed-effect (`lme()`) and nonlinear mixed-effect (`nlme()`) functions in the `nlme` package also accept a `weights`/`varIdent` argument of this type, to specify R-side (residual) heteroscedasticity.

## Hard example: factor-specific random-effects variance

### Example 1

Now consider the two-level model
$$
\begin{split}
Y_{ijk} & = \delta_i + b_{ij} + \epsilon_{ijk} \\
b_{ij} & \sim \text{Normal}(0,\sigma^2_{b_j}) \\
\epsilon_{ijk} & \sim \text{Normal}(0,\sigma^2_0) ;
\end{split}
$$
that is, now the structured variance is at the level of the among-group variance.
This question is asked in [http://thread.gmane.org/gmane.comp.lang.r.lme4.devel/7107](this r-sig-mixed-models thread), although it's not 100% clear whether the question really refers to scenario #1 above.


```{r simex2}
set.seed(101)
dsd <- c(1,2,3)
nblock <- 25
ntot <- 750
nfac <- 3
d <- expand.grid(f=letters[1:nfac],g=1:nblock,rep=seq(ntot/(nblock*nfac)))
d$fg <- with(d,interaction(g,f))
delta <- 4:6
## have to be a bit careful to get everything in the right order here ...
u <- rnorm(nblock*nfac,mean=0,sd=rep(dsd,each=nblock))
d$y <- rnorm(nrow(d),mean=delta[d$f],sd=0.2)+u[d$fg]
qplot(fg,y,geom="boxplot",colour=f,data=d)+coord_flip()
```

This *can* be done in `lme`: badly, as follows:
```{r badlme}
lme(y~f-1,random=~f|g,data=d)
```

Or correctly:
```{r correctlme}
lme(y~f-1,random=list(g=pdDiag(~f)),data=d)
```

It can only be done in `lme4` the bad way (at present).

```{r badlme4,message=FALSE}
detach("package:nlme",unload=TRUE)
library("lme4")
lmer(y~f-1+(f|g),data=d)
```

We get approximately the right standard deviation values, but with bogus correlations (the off-diagonal elements must be zero in this case because of the way we defined the model ...)

W. Volterman points out that we can get the right variance components, although with some bogus information thrown in, by specifying an interaction of the nested blocks with the higher-level groups, as follows: 

```{r kluge1,eval=FALSE}
lmer(y~f-1+(fg|g),data=d)
```

However, for this example the model takes too long to run (see below for an example that does work, and a better way to do this).


### Example 2

A much smaller example proposed by WV:

```{r wvex}
q3 <- structure(list(score = c(2L, 5L, 2L, 7L, 9L, 8L, 4L, 3L, 6L, 
4L, 2L, 6L, 10L, 8L, 9L, 8L, 6L, 2L, 3L, 1L, 1L, 3L, 2L, 8L, 
6L), course = structure(c(2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 
2L, 3L, 3L, 3L, 3L, 3L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L
), .Label = c("C", "E", "G"), class = "factor"), section = structure(c(5L, 
5L, 5L, 6L, 6L, 7L, 7L, 7L, 7L, 7L, 8L, 8L, 9L, 9L, 9L, 1L, 2L, 
2L, 2L, 2L, 3L, 3L, 3L, 4L, 4L), .Label = c("C1", "C2", "C3", 
"C4", "E1", "E2", "E3", "G1", "G2"), class = "factor")), .Names = c("score", 
"course", "section"), class = "data.frame", row.names = c(NA, 
-25L))
```

```{r}
library(lme4)
q3.lme4<-lmer(score~0+course+(0+course|section),data=q3)
summary(q3.lme4)
(x1<-ranef(q3.lme4))
```

Plots include estimates which should be zero (but aren't: they could be eliminated, with enough hacking around).

```{r}
dotplot(ranef(q3.lme4,postVar=TRUE))
```

```{r}
qqmath(ranef(q3.lme4,postVar=TRUE))
```

We can fit the same model with `lme()` 
(this is not actually run, because of [an issue with development lme4 and 
nlme loading/unloading](https://github.com/lme4/lme4/issues/35) ...)
```{r lmefit,eval=FALSE}
detach("package:lme4", unload = TRUE)
library(nlme)
q3.nlme<-lme(score~0+course,random=~0+course|section,data=q3)
summary(q3.nlme)
x2<-ranef(q3.nlme)
detach("package:nlme", unload = TRUE)
```

```{r x2hack,echo=FALSE}
x2 <- structure(list(courseC = c(2.16396655012683, -1.57514612672732, 
-2.3784392834401, 1.78961886004059, -3.80141213550967e-06, 4.25330747161918e-06, 
-4.51895336109504e-07, -9.53415751426565e-06, 9.53415751426564e-06
), courseE = c(3.15794393084079e-06, -2.29865990818173e-06, -3.47093068489567e-06, 
2.61164666198944e-06, -1.81868455612278, 2.03488186372044, -0.216197307597661, 
-1.54733896638393e-05, 1.54733896638393e-05), courseG = c(1.54966368952568e-05, 
-1.12799652942525e-05, -1.70325229616974e-05, 1.28158513608922e-05, 
-3.02747899662938e-05, 3.38737252827087e-05, -3.59893531663192e-06, 
-2.2265349503957, 2.2265349503957)), .Names = c("courseC", "courseE", 
"courseG"), row.names = c("C1", "C2", "C3", "C4", "E1", "E2", 
"E3", "G1", "G2"), class = c("ranef.lme", "data.frame"), effectNames = c("courseC", 
"courseE", "courseG"), label = "Random effects", level = 1L, standardized = FALSE, grpNames = "section")
```

Estimated random effects are roughly the same (at least, the ones with any meaning are)
```{r compareRE}
colGrp <- substr(names(x2),7,7)
rowGrp <- substr(rownames(x2),1,1)
x2mat <- as.matrix(x2)
x2mat[rowGrp[row(x2)]!=colGrp[col(x2)]] <- NA
x2 <- as.data.frame(x2mat)
all(abs(na.omit(unlist(x1[[1]]-x2)))<4e-6)
```

We can do better with the development version of `lme4` and some hacking, 
fitting by wrapping deviance function, setting relevant $\theta$ parameters to zero.
In the long run it would be better (more efficient) to change `Lind`/`Lambdat` in `mkReTrms`.  

If the lower triangle of the Cholesky factor of a $2 \times 2$ matrix is $\{\theta_1,\theta_2,\theta_3\}$ then the matrix is
$$
\left(
\begin{array}{cc}
\theta_1^2 & \theta_1 \theta_2 \\
\theta_1 \theta_2 & \theta_2^2 + \theta_3^2
\end{array}
\right)
$$
So if we set $\theta_2=0$ we get a diagonal matrix with variances $\theta_1^2$,
$\theta_3^2$.

### Implementation

See `help("modular",package="lme4")` for a discussion of the modular steps
of `lmer`-fitting ...

```{r diagel,echo=FALSE}
## indices of diagonal elements
## (not currently used, maybe handy in the future)
diag_el <- function(n) {
  m <- matrix(1:n^2,nrow=n)
  which(m[lower.tri(m,diag=TRUE)] %in% diag(m))
}
```

The first two modular steps process the formula and construct a deviance function:
```{r}
library("lme4")
lmod <- lFormula(score~0+course+(0+course|section),data=q3)
devfun <- do.call(mkLmerDevfun, lmod)
```

Now we need a wrapper function that will take a vector of diagonal elements
and place them into the right positions to construct a Cholesky factor with
the off-diagonal elements equal to zero ... we use the `lower` component
stored in the environment of the function (which is copied from the original
deviance function) to figure out how long the vector should be and where
the diagonal elements fall (these are identified as the elements with lower
bounds equal to zero).
```{r}
devfunw <- function(theta) {
  n <- length(lower)  ## from environment
  th <- numeric(n)
  diag_el <- which(lower==0)
  th[diag_el] <- theta
  devfun(th)
}
environment(devfunw) <- environment(devfun)
```

Now we test the function with simple inputs ($\theta_1=\theta_2=\theta_3=1$);
run the function in with `Nelder_Mead()`;
and use `mkMerMod` to construct the final output.  (I would like to use
the built-in `optimizeLmer()` function for the optimization step, but
there are a few little glitches there: among other things, it would be nice
if `optimizeLmer()` took a `start` argument!)
```{r}
devfunw(c(1,1,1)) ## test
opt <- Nelder_Mead(devfunw,par=c(1,1,1))
(res <- mkMerMod(environment(devfun), opt, lmod$reTrms, fr = lmod$fr))
```

Build a more robust/complete function.  It assumes there is
a single RE grouping factor and that all off-diagonal elements of the
Cholesky factor (corresponding to all off-diagonal elements of the
variance-covariance matrix of the RE) are zero.
```{r diagLmerFun}
diagLmer <- function(formula,data) {
   lmod <- lFormula(formula,data)
   devfun <- do.call(mkLmerDevfun, lmod)
   lower <- environment(devfun)$lower
   n <- length(lower)
   n0 <- (-1 + sqrt(1+8*n))/2  ## number of diagonal elements
   devfunw <- function(theta) {
     th <- numeric(n)
     diag_el <- which(lower==0)
     th[diag_el] <- theta
     devfun(th)
   }
   opt <- Nelder_Mead(devfunw,par=rep(1,n0),
                      lower=rep(0,n0))
   mkMerMod(environment(devfun), opt, lmod$reTrms, fr = lmod$fr)
}
```

```{r diagfit1}
print(diagLmer(score~0+course+(0+course|section),data=q3),cor=FALSE)
```

```{r diagfit2}
fit2 <- diagLmer(y~f-1+(f|g),data=d)
print(fit2,cor=FALSE)
```


## Autogregressive models

This should probably be broken out into a separate document.  I'm excited about
it because it suggests a natural way to set up a multilevel GLMM with temporal
autocorrelation:

* create a factor `f.obs` with a separate value for each level
* cross this factor with a *single-level* grouping factor `one` so that the random effects represent a *single* draw from a MVN distribution with an AR1 variance-covariance matrix;
* the parameters of the RE model are $\sigma^2$, $\rho$.  The Cholesky parameters $\theta_{i=1 \dots n}$ (the first column) are $\sigma \rho^{i-1}$; the Cholesky parameters beyond the first column in the $j^\text{th}$ off-diagonal (where $j=1$ represents the diagonal) are $\sigma \rho^{j-1} \sqrt{1-\rho^2}$.

The variance-covariance matrix for $n=3$ is:
$$
\left(
\begin{array}{ccx}
\theta_1^2 & \theta_1 \theta_2  & \theta_1 \theta_3 \\
\theta_1 \theta_2 & \theta_2^2 + \theta_4^2 & \theta_2 \theta_3 + \theta_4 \theta_5 \\
\theta_1 \theta_3 & \theta_2 \theta_3 + \theta_4 \theta_5 & \theta_3^2 + \theta_5^2  + \theta_6^2
\end{array}
\right)
$$

For an AR1 model, by hand, this gives a Cholesky factor of:
$$
\left(
\begin{array}{ccx}
1 & 0 & 0  \\
\rho & \sqrt{1-\rho^2} & 0 \\
\rho^2 & \rho \sqrt{1-\rho^2 } & \rho \sqrt{1-\rho^2}
\end{array}
\right)
$$

For what it's worth, here's the C code from the `nlme` package (`src/corStruct.c`) that returns the "transpose inverse square root factor" of an AR1 matrix (see `?corFactor.corStruct`)
```
static void
AR1_fact(double *par, longint *n, double *mat, double *logdet)
{
    longint i, np1 = *n + 1;
    double aux = sqrt(1 - *par * (*par)), aux1 = - (*par)/aux;

    *logdet -= (*n - 1) * log(aux);
    aux = 1/aux;
    mat[0] = 1;
    for(i = 1; i < *n; i++) {
        mat[i * np1] = aux;
        mat[i + *n * (i - 1)] = aux1;
    }
}
```

```{r ar1fact}
ar1_chol <- function(par,n) {
  mat <- matrix(0,nrow=n,ncol=n)
  mat[1] <- 1
  aux <- sqrt(1-par^2)
  np1 <- n+1
  aux1 <- -par/aux
  aux <- 1/aux
  for (i in 1:(n-1)) {
    mat[i*np1+1] <- aux
    mat[i + n*(i-1)+1] <- aux1
  }
  mat
}
(a1 <- ar1_chol(0.5,3))
a1 %*% t(a1)
```

```{r}
library("nlme")
n <- 3
ii <- Initialize(corAR1(0.5),data=data.frame(x=1:n))
(mm <- matrix(corFactor(ii),nrow=n)) ## or corMatrix(ii,corr=FALSE)
ss <- solve(mm)
ss %*% t(ss)
chol(ss %*% t(ss))
n <- 9
ii2 <- Initialize(corAR1(0.5),data=data.frame(x=1:n))
chol(corMatrix(ii2))
```

By inspection, the first row/column is $\rho^{i-1}$; the diagonals leading down therefrom are $\rho^{i-1} \sqrt{1-\rho^2}$ ...

Want the inverse-Cholesky factor for applying to residuals; want
the Cholesky factor for applying to $\theta$.

## Thoughts on syntax, design, etc..

(More stuff that should really be separate)

Classes of variance-covariance matrices:

* identity (`MCMCglmm`: `cor`)
* constant variance (`MCMCglmm`: `idv`, `nlme`: `pdIdent`)
* diagonal (`MCMCglmm`: `idh`, `nlme`: `pdDiag`)
* compound symmetry (`nlme`: `corCompSymm`)
* unstructured (`MCMCglmm`: `us`, `nlme`: `pdSymm`)
* blocked (`nlme`: `pdBlocked`)
* *time series*: AR1 etc.
* *spatial*: Gaussian etc.
* *phylogenetic* and *pedigree*: Brownian etc.
* row/column, Toeplitz

Would be worth look at [http://www.vsni.co.uk/software/asreml/htmlhelp/asreml/vcode.htm](ASREML's list of variance structures) for comparison.

When designing this it would be worth following the `nlme` design to some extent, creating objects of some sort that, for specified parameters and dimensions, can return (1) correlation (or V-C) matrices [for reporting]; (2) Cholesky factors [for forward computation, when constructing $L$ matrices for RE computation for G-side effects]; (3) inverse Cholesky factors [for computation of R-side effects], upon request.  If possible it would be nice to make the design a little more transparent than `nlme`'s ...
