## Convergence testing

```{r pkgs,message=FALSE}
library(lattice)
library(lme4)
library(blme)
library(reshape2)
library(ggplot2); theme_set(theme_bw())
library(gridExtra)  ## for grid.arrange
library(bbmle) ## for slice2D; requires *latest* r-forge version (r121)
source("allFit.R")
```

Load data:

```{r getdat}
dList <- load("data.RData")
```

Spaghetti plot: don't see much pattern other than (1) general
increasing trend; (2) quantized response values (`table(dth$Estimate)`
or `unique(dth$Estimate)` also show this); (3) skewed residuals

```{r usort}
sort(unique(dth$Estimate))
```

```{r spaghett1,warning=FALSE}
(p0 <- ggplot(dth,aes(Actual,Estimate))+geom_point()+
    geom_line(aes(group=factor(pid)))+
    stat_summary(fun.y=mean,geom="line",colour="red",lwd=2))
```
Emphasize distribution:
```{r sumplot}
ggplot(dth,aes(x=Actual,y=Estimate))+stat_sum(aes(size=..n..),alpha=0.5)+
    scale_size_continuous(range=c(2,8))
```

```{r fit1,message=FALSE,results="hide"}
getcor <- function(object) attr(VarCorr(object)[[1]],"correlation")[2,1]
f1 <- lmer(Estimate~Actual+(Actual|pid),data=dth,REML=FALSE)
getcor(f1)
```

logging (there are zero values: it's a bit sloppy, but I'm just
going to use `log(0.5+x)` as the response -- `log(1+x)` may look
even better ...) helps a great
deal with the fit (no longer singular), and with the Q-Q

```{r sumplot2}
ggplot(dth,aes(x=Actual,y=log(0.5+Estimate)))+stat_sum(aes(size=..n..),alpha=0.5)+
    scale_size_continuous(range=c(2,8))+
    geom_smooth(method="lm")
```

```{r fit1L}
f1L <- lmer(log10(0.5+Estimate)~Actual+(Actual|pid),data=dth,REML=FALSE)
VarCorr(f1L)
```

Q-Q plot shows the problem ...

```{r qq,fig.width=10}
grid.arrange(qqmath(f1),qqmath(f1L),nrow=1)
```

and helps the general residual plot

```{r diag,fig.width=10}
grid.arrange(plot(f1,type=c("p","smooth")),
             plot(f1L,type=c("p","smooth")),nrow=1)
```

Double-check that the same thing happens with `lme4.0`

```{r otherfits}
f2 <- lme4.0::lmer(Estimate~Actual+(Actual|pid),data=dth,REML=FALSE)
attr(lme4.0::VarCorr(f2)[[1]],"correlation")[2,1]

```{r allfits,results="hide",warning=FALSE,cache=TRUE}
aa <- allFit(f1)
ss <- summary.allfit(aa)
```

results: warning for L-BFGS-B (which does indeed fail); all other
results nearly identical (`nlminb` gives cor *slightly* <1) (only
sd/cor estimates shown; log-likelihoods and fixed effect estimates
are consistent with this picture)
(Facet and axis tick labels below are slightly wonky/need tweaking.)

```{r sdcor1,message=FALSE,warning=FALSE,fig.width=10}
names(which(!sapply(ss$msgs,is.null)))
(g0 <- ggplot(melt(ss$sdcor),aes(Var2,value,colour=Var1))+
    geom_point(position=position_dodge(width=0.5))+
    facet_wrap(~Var2,scale="free"))
```

## blmer
L-BFGS-B fails completely (unusual error, missing value pops up in internal boundary test -- should investigate)
none of the other cases gives convergence warnings, which is slightly alarming since both versions of BOBYQA (`minqa` and `nloptr`) clearly fail, with NLL 50-100 log-likelihood units worse than the other fits ... I think the problem here is actually something in `blmer` where the warnings fail to get stored properly ...


```{r blmefit}
## hack to refit lmer as glmer
ff <- getCall(f1)
ff[[1]] <- quote(blmer)
f3 <- eval(ff)
aa3 <- allFit(f3)
ss3 <- summary.allfit(aa3)
```

Despite the failure of `blme` to *store* the warnings, we can
see from the output that the warnings do correspond to the
incorrect results (i.e., both BOBYQA implementations).
It's interesting that BOBYQA fails here, since in most
other cases we have found it to be the *most* robust algorithm ...

```{r blmeplot,message=FALSE,fig.width=10}
g0 %+% melt(cbind(ss3$sdcor,NLL=-2*(ss3$ll-max(ss3$ll))))
```

## Visualizing likelihood surface


```{r slicecalc,cache=TRUE,results="hide"}
rmat <- matrix(c(0,1,-0.2,0.2,0,1),byrow=TRUE,ncol=2,
               dimnames=list(paste0("theta",1:3),c("lwr","upr")))
dd1 <- update(f1,devFunOnly=TRUE)
pp1 <- getME(f1,"theta")
s2D.1 <- bbmle:::slice2D(pp1,dd1,
                       tranges=rmat)
##
dd3 <- update(f3,devFunOnly=TRUE)
pp3 <- getME(aa3[[2]],"theta")  ## use a version that actually worked
s2D.3 <- bbmle:::slice2D(pp3,dd3,
                       tranges=rmat)
```

Left plot, `lmer`; right, `blmer`.


```{r sploms,fig.width=10}
grid.arrange(splom(s2D.1),splom(s2D.3),nrow=1)
```

Zoom in a bit:

```{r slicecalc2,cache=TRUE,results="hide"}
rmat2 <- matrix(c(0,1,-0.1,0.1,0,0.1),byrow=TRUE,ncol=2,
               dimnames=list(paste0("theta",1:3),c("lwr","upr")))
s2D.1b <- bbmle:::slice2D(pp1,dd1,
                       tranges=rmat2)
s2D.3b <- bbmle:::slice2D(pp3,dd3,
                       tranges=rmat2)

```{r sploms2,fig.width=10}
grid.arrange(splom(s2D.1b),splom(s2D.3b),nrow=1)
```

Visualizations could use a lot of work (scale bar and/or
contours; allow for more points, e.g. visualize where non-converging
fits ended up ...)

### Gauss with log link?

This is a nice idea but fails ... if we don't
tweak the LHS we get `NaN` values immediately
(these *should* provoke a more useful warning/error message!);
if we use an offset of 0.1 then we get a PWRSS failure; with
an offset of 0.5 we get another singular fit.
```{r}
(f4 <- glmer(Estimate+0.5~Actual+(Actual|pid),data=dth,
            family=gaussian(link="log")))
```

Stuff left out (so far):

* other data sets (`dord`, `dorltc`) (could follow similar strategies)
