Mixed model simulations
========================================================

What happens if we try fit mixed models with too few samples? Here's a little example.

Let's consider just two-level mixed (actually pure random-effect) models, for simplicity.

A function to simulate data and one to fit the model and extract the estimated random-effect standard deviation.  (This is done with development `lme4` but the results would probably be nearly identical in stable `lme4` or in `nlme`)

```{r message=FALSE}
simfun <- function(n1=5,n2=5,sd1=1,sd2=1) {
  d <- expand.grid(f1=factor(seq(n1)),f2=factor(seq(n2)))
  u1 <- rnorm(n1,sd=sd1)
  d$y <- rnorm(n1*n2,mean=u1,sd=sd2)
  d
}
require(lme4)
fitfun <- function(d=simfun()) {
  sqrt(unlist(VarCorr(lmer(y~(1|f1),data=d))))
}
```

Run 500 sims with `n1=5` samples at the top level and 500 with `n1=3`

```{r}
set.seed(101)
```

```{r runsims,cache=TRUE}
sd_dist1 <- replicate(500,fitfun())
sd_dist2 <- replicate(500,fitfun(simfun(n1=3)))
sd_List <- list(n1.5=sd_dist1,n1.3=sd_dist2)
```

Histograms with slightly-prettier-than-default settings:
```{r}
plotfun <- function(x,trueval) {
  par(las=1,bty="l")
  hist(x,breaks=50,col="gray",main="",xlab="est. sd",
       freq=FALSE)
}
```

Plot results:
```{r plots,fig.width=7,fig.height=4}
par(mfrow=c(1,2))
invisible(lapply(sd_List,plotfun))
```
We can see that there is a spike at zero in both cases, although it is much larger when `n1=3`.
Proportion of samples *exactly* equal to zero:

```{r}
sapply(sd_List,function(x) mean(x==0))
```

If we were lucky (in some sense), these zeros would
be counterbalanced by large values so that we could
say the overall estimate was unbiased (i.e., the
*mean* estimate was equal to the true value, even
if the estimates varied all over the place).

This isn't true though.  Here we compute the mean,
standard error, and confidence intervals ($\pm 2 \text{SEM}$) for both runs:

```{r}
sfun <- function(x)  {
  r <- list(mean=mean(x), sem=sd(x)/sqrt(length(x)))
  r <- with(r,c(r,list(lwr=mean-2*sem,upr=mean+2*sem)))
  unlist(r)
}
print(s_tab <- sapply(sd_List,sfun),digits=3)
bias_pct <- round((1-s_tab["mean",])*100)
```

`n1=5` is biased low by `r bias_pct["n1.5"]`%;
`n1=3` is biased low by `r bias_pct["n1.3"]`%.

And by the way, this is with the default `REML=TRUE` setting.  I would also guess that this is generic behaviour of modern mixed models, not a problem with R's code -- I would guess you'd get the same sorts of results from ASREML/Genstat/SAS PROC MIXED etc.
