Salamander GLMM
==========================

```{r opts,echo=FALSE}
opts_chunk$set(tidy=FALSE,fig.width=6,fig.height=4)
```

```{r pkgs,message=FALSE,warning=FALSE}
library(plyr)
library(lme4)
library(glmmADMB)
library(bernor)
library(MASS)
library(nlme)
library(hglm)
```

```{r utils,echo=FALSE}
sumfun <- function(m) c(fixef(m),unlist(VarCorr(m)))
```

The "salamander mating" data set is a classic example in generalized linear mixed models (e.g. see this example [from Charlie Geyer](http://www.stat.umn.edu/geyer/bernor/example7.pdf), or this one [from the SAS PROC GLIMMIX documentation](http://support.sas.com/documentation/cdl/en/statug/63962/HTML/default/viewer.htm#statug_glimmix_sect016.htm)).  Geyer points out that this example has been discussed by:

>  McCullagh and Nelder (1989), Schall (1991), Karim and Zeger
(1992), Breslow and Clayton (1993), Wolfinger and O’Connell (1993), and Shun
(1997).

(More recently, it has been analyzed by @booth_maximizing_1999 and by @sung_monte_2007.)

Estimates of female and male variance parameters from
SAS PROC GLIMMIX:
```{r SASests,echo=FALSE}
(SASests <- matrix(c(1.4099,0.8871,0.08963,0.4102),
                  byrow=TRUE,nrow=2,
                  dimnames=list(c("fnum","mnum"),
                                c("var_est","var_se"))))
```

Estimates of all parameters (from @booth_maximizing_1999 by
way of Geyer's example).
```{r BHests,echo=FALSE}
(BHests <- c("R/R"=1.030,"R/W"=0.320,
             "W/R"=-1.950,"W/W"=0.990,
             "sigma_f"=1.183 ,"sigma_m"=1.118))
```
Geyer considers these results to be the gold standard (he gets close to them with a sufficiently thorough Monte Carlo maximum likelihood fit, and says he has independently confirmed with an MCMC fit).
As shown below, it seems as though most available algorithms
estimate the among-female variance by quite a lot (orders of
magnitude, although surprisingly (to me) SAS PROC GLIMMIX is
one of the closest ...

```{r getdat}
sdat <- read.table("salamander.txt",header=TRUE,
                   colClasses=c(rep("factor",5),"numeric"))
```

Check experimental design:
```{r exptabs}
with(sdat,table(fpop,mpop))
with(sdat,table(fnum,mnum))
```

## Methods

### lme4

```{r lme4fit,message=FALSE}
library(lme4)
LM_t_0 <- system.time(
  LM_fit_0 <- glmer(mating~0+fpop:mpop+
        (1|mnum)+(1|fnum),
      sdat,family=binomial))
  
print(LM_fit_0,cor=FALSE)
```

### glmmADMB
```{r GA0,message=FALSE}
library(glmmADMB)
GA_t_0 <-
  system.time(
    GA_fit_0 <- glmmadmb(mating~0+fpop:mpop+
        (1|mnum)+(1|fnum),
      sdat,family="binomial"))
```

#### Importance sampling

Using importance sampling to try to do better (see [D. Fournier's post](https://groups.google.com/forum/#!topic/admb-users/7RltSt9Pm1A)):
```{r GA_is, cache=TRUE}
## 25 points, 50 samples, random number seed=121:
GA_t_is1 <- system.time(
  GA_fit_is1 <- glmmadmb(mating~0+fpop:mpop+
        (1|mnum)+(1|fnum),
      sdat,family="binomial",
          extra.args="-is 25  121  50"))
GA_t_is2 <- system.time(
  GA_fit_is2 <- glmmadmb(mating~0+fpop:mpop+
        (1|mnum)+(1|fnum),
      sdat,family="binomial",
          extra.args="-is 50  121  100"))
```

```{r cmat,echo=FALSE}
nvec <- c("glmer","GA_0","GA_is_50","GA_is_100")
mList <- list(LM_fit_0,GA_fit_0,GA_fit_is1,GA_fit_is2)
cmat <- t(laply(mList,sumfun))
colnames(cmat) <- nvec
print(cmat,digits=3)
```
*Conclusion*: Results are similar from both approaches; Laplace approximation doesn't work that well. Importance sampling only helps a bit.

#### glmmADMB with MCMC

```{r GA_mcmc,cache=TRUE}
GA_t_mcmc <- system.time(
  GA_fit_mcmc <- glmmadmb(mating~0+fpop:mpop+
        (1|mnum)+(1|fnum),
      sdat,family="binomial",
      mcmc=TRUE))
```

Diagnose (the `tmpL` variables represent the
variance parameters, on a standard deviation scale).
```{r traceplot}
library(coda)
xyplot(GA_fit_mcmc$mcmc[,1:6],layout=c(2,3),aspect="fill")
```

A close-up of the female-sd parameter:

```{r sdfhist,echo=FALSE}
par(bty="l",las=1)
hist(log10(GA_fit_mcmc$mcmc[,"tmpL.2"]),col="gray",breaks=50,
     main="",ylab="log(sd(F))",freq=FALSE)
```

Bottom line, these chains aren't mixing very well. We would have to run it for a lot longer, or fuss with tuning parameters like `mcgrope`, to get a good answer this way.

### glmmPQL

glmmPQL should in principle work, but
[various people have tried and failed](http://tolstoy.newcastle.edu.au/R/help/05/01/10005.html)
```{r glmmPQLfit,message=FALSE}
library(MASS)
library(nlme)
## http://tolstoy.newcastle.edu.au/R/help/05/01/10005.html
sdat$dummy <- factor(rep(1,nrow(sdat)))
sdatG <- groupedData(mating~1|dummy,sdat)
glmmPQL(mating~0+fpop:mpop,
        data=sdatG,
  random=pdBlocked(list(pdIdent(~mnum-1),
                        pdIdent(~fnum-1))),
        family=binomial)
## this one works, but only with groupedData
lme(mating~0+fpop:mpop,data=sdatG,
        random=pdBlocked(list(pdIdent(~mnum-1),pdIdent(~fnum-1))))
data(Assay)  ## from PB p 165
Assay <- as.data.frame(Assay)
fm1Assay <- lme(logDens~sample*dilut,Assay,
         random=pdBlocked(list(pdIdent(~1),pdIdent(~dilut-1),pdIdent(~sample-1))))
```
I think the proximal problem is that `glmmPQL` constructs a model frame from the data, which loses the `groupedData` class (and the `formula` attribute, which seems necessary -- I'm not sure why).  It might be hackable ...

### `bernor` package

Geyer and Sung's `bernor` package is probably the best available (=most accurate) R choice at the moment.  However, it needs some hacking to make it run with modern R. I have built `bernor` v. 0.3-8, can make it available if necessary.

The following code is reproduced from [Geyer's example](http://www.stat.umn.edu/geyer/bernor/example7.pdf) (this also demonstrates that Geyer's package, while functional, does not package everything nicely for the user as we have come to expect)

```{r bernor1}
set.seed(101)  ## initialize .Random.seed
objfun <- function(theta) {
   if (!is.numeric(theta))
   stop("objfun: theta not numeric")
   if (length(theta) != nparm)
      stop("objfun: theta wrong length")
   mu <- theta[seq(1, nfix)]
   sigma <- theta[-seq(1, nfix)]
   .Random.seed <<- .save.Random.seed
   bnlogl(y, mu, sigma, nmiss, x, z, i, moo)$value
}
objgrd <- function(theta) {
   if (!is.numeric(theta)) 
     stop("objfun: theta not numeric")
   if (length(theta) != nparm)
     stop("objfun: theta wrong length")
   mu <- theta[seq(1, nfix)]
   sigma <- theta[-seq(1, nfix)]
   .Random.seed <<- .save.Random.seed
  bnlogl(y, mu, sigma, nmiss, x, z, i, moo, deriv = 1)$gradient
}
## FIXME: should memoize!
library(bernor)
data(salam)
attach(salam)  ## would prefer not to do this ...
nparm <- ncol(x) + length(unique(i))
nfix <- ncol(x)
moo <- model("gaussian", length(i), 1)
.save.Random.seed <- .Random.seed
nobs <- ncol(y)
nmiss <- 100
theta.start <- rep(0, nparm)
names(theta.start) <- c(dimnames(x)[[2]], paste("sigma", c("f",
"m"), sep = "_"))
lower <- rep(0, nparm)
lower[1:ncol(x)] <- (-Inf)
trust <- 1
lowert <- pmax(lower, theta.start - trust)
uppert <- theta.start + trust
control <- list(fnscale = -10)
tout <- system.time(out <- optim(theta.start, objfun, objgrd,
method = "L-BFGS-B", lower = lowert, upper = uppert, control = control))
cat("elapsed time", tout[1], "seconds\n")
```

Retry with larger trust region (slow)
```{r bernor2}
nmiss <- 10000
theta.start <- out$par
lowert <- pmax(lower, theta.start - trust)
uppert <- theta.start + trust
control <- list(fnscale = signif(out$value, 1))
```

```{r trust2,cache=TRUE}
tout <- system.time(out <- optim(theta.start, objfun, objgrd,
  method = "L-BFGS-B", lower = lowert, upper = uppert, control = control))
cat("elapsed time", tout[1], "seconds\n")
```

Read the linked example by Geyer for more details: they go on to try `nmiss=1e7` (number of samples of "missing data" [=latent variables]?, which they say is very slow, but they also say their results match @booth_maximizing_1999 , and an independent MCMC estimate, to 3 significant digits).

### hglm


```{r hglm}
hglm_fit_0 <- hglm2(meanmodel = mating ~ 0 + fpop:mpop  + 
                       (1 | as.numeric(fnum)) + (1 | as.numeric(mnum)), 
                    family = binomial(), data = sdat,
                    conv = 1e-08, maxit = 40)
```

The example from the `hglm` vignette:
```{r}
data(salamander)
hglm.salam <- hglm2(meanmodel = Mate ~ TypeF + TypeM + TypeF*TypeM + (1 | Female) + (1 | Male), family = binomial(), data = salamander,conv = 1e-08, maxit = 40)
```

Note `hglm` has a different version of the salamander data set -- 360 rather than 120 values ... ('summer' data set is what we've entered above, I think, this is both summer and fall ... ...)
(Even for the data as included with `hglm` I don't seem to be getting exactly the same structure as in the `hglm` vignette ... ???)

## To do/Other possibilities
* SAS, AS-REML
* INLA?
* WinBUGS/JAGS/Stan?
* print package numbers etc.
* pictures of estimates, timings, confidence intervals ...
