Power/assessment of Poisson GLMMs
========================================================

Martí Casals writes:

> The final data employed for the analysis were composed of 2187 cases (27
> players * 81 games), having followed a filtering process. I dealt with a
> balanced study design with repeated measurements given that each player was
> observed the same number of games, and therefore the player was considered
> as a random effect. We carried out two mixed models, a Linear Mixed Model
> (LMM) through the nlme library (Pinheiro) and a Poisson mixed model with
> lme4 package (Bates).

```{r prelims}
library(lme4.0)  ## development version ...
library(plyr)
library(ggplot2)
theme_set(theme_bw())
library(reshape2)
set.seed(101)
```

```{r simdatafun}
simdata <- function(nplayers=27,ngames=81,sd.game=1,sd.player=1,sd.obs=0.5,
                    beta0=0.5) {
  d <- expand.grid(player=factor(seq(nplayers)),game=factor(seq(ngames)))
  u.game <- rnorm(ngames,sd=sd.game)
  u.player <- rnorm(nplayers,sd=sd.player)
  u.obs <- rnorm(ngames*nplayers,sd=sd.obs)
  d <- transform(d,obs=factor(seq(nrow(d))),
                   eta=beta0+u.game[game]+u.player[player]+u.obs)
  d$y <- rpois(nrow(d),exp(d$eta))
  d
}
```

```{r simdata1}
d1 <- simdata()
fit1 <- glmer(y~(1|player)+(1|game)+(1|obs),data=d1,family=poisson)
unlist(VarCorr(fit1))
simfun <- function(m=fit1,...) {
  sqrt(unlist(VarCorr(refit(m,simdata(...)$y))))
}
s0 <- simfun()
```

```{r simbatch,cache=TRUE}
nsim <- 500
simres <- raply(nsim,simfun()) 
```

```{r summary}
summary(simres)
```

```{r plot1}
library(proto)
## source("geom-hboxplot.r")
m <- transform(melt(simres),Var2=factor(Var2,levels=c("game","player","obs")))
truevals <- data.frame(Var2=c("game","obs","player"),value=c(1,0.5,1))
ggplot(m, aes(x=Var2,y=value))+geom_boxplot(notch=TRUE)+
     facet_wrap(~Var2,scale="free")+
  geom_hline(aes(yintercept=value),data=truevals,lwd=2,alpha=0.5,colour="red")+
  geom_violin(alpha=0.2,colour=NA,fill="blue")
```

This looks reasonably good on the whole, although the estimates are *slightly* biased (i.e. notches don't include true values for `player` and `obs`); this could
be a finite-size/REML-y issue?

Would be nice to turn the whole thing sideways, but: `ggplot2 does not currently support free scales with a non-cartesian coord or coord_flip.`