## Calculating the variance-covariance matrix of variance-covariance parameters

One relatively frequent request from `lme4` users is for the capability to compute the variance-covariance matrix of the variance parameters themselves.  In the `nlme` package this was (implicitly) available via the `intervals()` function in the `nlme` package (which returned Wald confidence intervals); `lme4` has switched to providing profile and parametric bootstrap, and *not* Wald confidence intervals, because the Wald intervals on variance-covariance parameters are often really, really bad.

Nevertheless, there are sometimes reasons that one wants to get the asymptotic (Wald) variance-covariance matrix of the variance-covariance parameters themselves.  This document shows how to do that for ML estimates, and discusses the issues more generally (so that hopefully a similar solution for the REML estimates will also be feasible soon). 

The basic problem is `lme4`'s deviance function uses a scaled Cholesky-factor parameterization: it's easy enough to get an approximate Hessian of these ($\theta$) parameters by finite differencing, but annoying to convert them to the $\sigma$ or $\sigma^2$ scale.  However, for ML estimates we can use an internal function (`devfun2`) that is parameterized on the $\sigma$ scale.  (To get asymptotic variance-covariance estimates for REML estimates we will either need to adapt `devfun2` to the REML scale, or figure out a more direct scale conversion: see notes below.

### Preliminaries

Fit the good old `sleepstudy` model:
```{r fitmodels,message=FALSE}
library(lme4)
fm1 <- lmer(Reaction ~ Days + (Days | Subject), sleepstudy)
```

For comparison with the likelihood profile (which uses ML rather than REML) we also fit the ML equivalent:
```{r mlfit}
fm1.ML <- refitML(fm1)
```

Also fit with `lme` so that we can compare the results of `intervals()`:
```{r lmefit,message=FALSE}
library(nlme)
fm1.lme <- lme(Reaction ~ Days, random = ~Days| Subject, sleepstudy)
fm1.lme.ML <- update(fm1.lme,method="ML")
``` 

### Finite differences on the $\sigma$ scale

The easiest way to do what we want is to the internal `devfun2()` function, which is used within `profile` --- it is essentially the same as the basic `devfun()`, but it uses a standard deviation/correlation parameterization (and adds the residual standard deviation as a parameter) rather than a Cholesky parameterization:
```{r devfun2}
dd.ML <- lme4:::devfun2(fm1.ML,useSc=TRUE,signames=FALSE)
```
At the moment I think `devfun2()` cannot be used for REML fits
(as it is designed for computing likelihood profiles, that wouldn't
make much sense).

The other way to do this would be to compute the Hessian for `devfun2`, then do the requisite calculus and linear algebra on the function ${\cal F}$ that translates from the Cholesky to the stddev/corr parameterization to convert the Hessian to the new parameterization.

Retrieve the standard deviation/correlation parameters, in the correct order:
```{r getpars}
vv <- as.data.frame(VarCorr(fm1.ML)) ## need ML estimates!
## name for consistency with profile object below
vnames <- c(sprintf(".sig%02d",1:3),".sigma")
pars <- setNames(vv[c(1,3,2,4),"sdcor"],vnames)
```

Compute second derivatives (the factor of 2 converts from the deviance to the log-likelihood scale):
```{r gethess}
library(numDeriv)
hh1 <- hessian(dd.ML,pars)
vv2 <- 2*solve(hh1)
```

### Results and comparisons

Now that we have the Wald variance-covariance matrix, I'm going
to cook up a little bit of structure that will let us use 
`confint.default` to compute the Wald intervals:

```{r confints}
dimnames(vv2) <- list(vnames,vnames)
vhack <- list(coefficients=pars,vcov=vv2)
vcov.default <- function(object,...) object$vcov
(wci <- confint.default(vhack))
```

These are close, although not precisely the same as, the Wald
intervals produced by `lme` (note that the order differs ---
the `lme4` stuff is using "lower-triangular" order, so that
the correlation (`.sig02`) comes between the two standard deviation estimates,
while `lme` (and the `as.data.frame.VarCorr.merMod` method from `lme4`)
puts the correlations last).

```{r intervals}
(lmeint <- intervals(fm1.lme.ML,which="var-cov"))
```

#### Comparison with likelihood profile

Compute the likelihood profile:
```{r fitprof,cache=TRUE}
pp <- profile(fm1,which="theta_")
```

The linear approximation to the profile for a focal parameter $x$ should
be $\zeta \approx (x-\hat x)/\sigma_x$, where $\hat x$ is the MLE and
$\sigma_x$ is the Wald standard deviation:
```{r profplot}
dfprof <- as.data.frame(pp)
mframe <- data.frame(.par=vnames,ctr=pars,slope=1/sqrt(diag(vv2)))
dfprof <- merge(dfprof,mframe)
library(ggplot2); theme_set(theme_bw())
ggplot(dfprof,aes(.focal,.zeta))+geom_line()+geom_point()+
    facet_wrap(~.par,scale="free")+
    geom_hline(yintercept=0,color="gray")+
    geom_hline(yintercept=c(-1.96,1.96),color="gray",lty=2)+
    geom_line(aes(y=(.focal-ctr)*slope),color="red")
```    
In this particular example, there's not a huge divergence from
linearity over the range defined by the 95% confidence intervals
(dashed horizontal lines) ... as confirmed by the confidence intervals.

#### Comparison of CIs

```{r cmpconf,echo=FALSE,fig.height=3,fig.width=7}
pci <- confint(pp)
nf  <- function(x) setNames(as.data.frame(x),c("lwr","upr"))
lme4Dat <- rbind(
   data.frame(method="profile_lme4",
              par=rownames(pci),est=pars,nf(pci)),
    data.frame(method="Wald_lme4",
               par=rownames(wci),est=pars,nf(wci)))
lmeDat <- data.frame(method="Wald_lme",
       par=vnames,est=pars, ## close enough to identical
     rbind(nf(lmeint$reStruct$Subject[c(1,3,2),c(1,3)]),
           setNames(lmeint$sigma[c(1,3)],c("lwr","upr"))))
allDat <- rbind(lme4Dat,lmeDat)
ggplot(allDat,aes(par,est,ymin=lwr,ymax=upr,colour=method))+
    geom_pointrange(position=position_dodge(width=0.5))+
    facet_wrap(~par,scale="free",nrow=1)
```
Hmm.  Perhaps the `lme` Wald intervals are calculated on a different
scale, e.g. the log-$\sigma$ scale?  They look even closer to the
profile CIs ...

### Continuing

Second derivative via chain rule?  We're looking for $\frac{\partial^2 f(g(x))}{\partial x^2}$ (or $\frac{\partial^2 (f \circ g)(x)}{\partial x^2}$ may be easier notation in this case).

I should have been able to figure this out myself, but looking at
[Faà di Bruno's formula](http://en.wikipedia.org/wiki/Fa%C3%A0_di_Bruno%27s_formula):

$$
\frac{\partial^n}{\partial x_1 \dots \partial x_n} =
\sum_{\pi \in \Pi} f^{(|\pi|}(y) \cdot \prod_{B \in \pi} \frac{\partial^{|B|}y}{\prod_{j \in B} \partial x_j}
$$

For $n=2$ the partitions are just $\{\{1\},\{2\}\}$ and $\{\{1,2\}\}$
$$
\frac{\partial^2 f(y)}{\partial x_1 x_2} = 
f'(y) \frac{\partial^2 y}{\partial x_1 x_2} + f''(y) \frac{\partial y}{\partial x_1} \frac{\partial y}{\partial x_2}
$$

The derivative of a matrix cross-product $\bm X \bm X^T$ should be fairly straightforward (I think?); e.g. for a two-by-two crossproduct works out as $(\bm\lambda \bm\lambda^T)' = \bm\lambda' \bm\lambda^T + \bm\lambda \bm(\lambda^T)'$, where differentiation is element-by-element on the RHS.
